---
title: >
  Matching a single-cell RNA-seq response to perturbation references using
  perturbMatch
author:
- name: Jamie Soul
  affiliation:
  - Biochemistry, Cell and Systems Biology, Institute of Systems, Molecular
    & Integrative Biology, Faculty of Health and Life Sciences
  email: jamie.soul@liverpool.ac.uk
date: "`r Sys.Date()`"
package: perturbMatch
output:
  BiocStyle::html_document:
    titlecaps: false
    toc: true
    toc_float: true
    number_sections: true
abstract: >
  perturbMatch scores a differential expression signature against a reference
  collection of single-gene perturbations, whatever the source of the
  signature. This vignette applies perturbMatch to a single-cell RNA-seq
  experiment: interferon-stimulated versus control CD14+ monocytes from the
  Kang et al. PBMC dataset. It covers aggregating cells into pseudobulk
  samples, scoring the resulting signature, recognising and filtering hits
  that cannot be genuine human regulators, and using singscore to project a
  candidate regulator's activity back onto individual cells.
vignette: >
  %\VignetteIndexEntry{perturbMatch single-cell}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
library(knitr)
opts_chunk$set(error = FALSE, message = FALSE, warning = FALSE)
```

```{r setup, message=FALSE, warning=FALSE}
library(perturbMatch)
library(muscData)
library(SingleCellExperiment)
library(scuttle)
library(DESeq2)
library(SummarizedExperiment)
library(singscore)
library(scater)
library(ggplot2)
```

# Overview

perturbMatch analyses a differential expression signature regardless of
whether it came from bulk or single-cell RNA-seq. For single-cell data the
recommended route is to aggregate counts into pseudobulk samples and run a
standard differential expression analysis. This vignette follows that route on the Kang et al. PBMC dataset
from `muscData`, comparing control and interferon-beta-stimulated CD14+
monocytes.

The bulk RNA-seq vignette (`vignette("perturbMatch")`) introduces the
reference database, scoring methods, and plotting functions in more detail.
Here we focus on the single-cell workflow and two points specific to it:
filtering hits that cannot be human regulators, and validating a candidate
regulator at the level of individual cells.

# Load the data

The `muscData` package distributes the Kang et al. PBMC study as a
`SingleCellExperiment`. Eight lupus patients each contributed peripheral
blood mononuclear cells that were split into two aliquots, one left untreated
and one stimulated with recombinant interferon beta for six hours, then
profiled by droplet single-cell RNA-seq. The design therefore pairs a control
and a stimulated sample within every donor, which lets us treat donor as a
blocking factor later on. The object carries raw UMI counts in the `counts`
assay, and per-cell annotation in `colData` including the donor of origin
(`ind`), the stimulation condition (`stim`), a coarse cell type label
(`cell`), and a doublet call (`multiplets`).

We copy the donor identifier into a tidier `donor` column and take a first
look at how many cells fall in each cell type and in each donor by condition
combination. The counts confirm the paired control and stimulated design
across the eight donors.

```{r load}
sce <- Kang18_8vs8()
sce$donor <- paste0("Donor", sce$ind)

table(sce$cell)
table(sce$donor, sce$stim)
```

# Select a cell population

We restrict the analysis to a single, well-defined population, the CD14+
monocytes, keeping only confidently assigned singlets.

```{r filter-cells}
sce <- sce[, sce$multiplets == "singlet"]
sce <- sce[, !is.na(sce$cell)]
sce <- sce[, sce$cell == "CD14+ Monocytes"]

sce
```

# Aggregate into pseudobulk samples

`aggregateAcrossCells()` sums counts within each donor and stimulation
condition, giving one pseudobulk profile per biological replicate.

```{r pseudobulk}
pb <- aggregateAcrossCells(
    sce,
    ids = DataFrame(donor = sce$donor, stim = sce$stim)
)

colData(pb)[, c("donor", "stim", "ncells")]
```

# Differential expression

We test for genes responding to interferon stimulation with DESeq2,
controlling for donor.

```{r de}
dds <- DESeqDataSet(pb, design = ~ donor + stim)
dds <- DESeq(dds)

res <- results(dds, contrast = c("stim", "stim", "ctrl"))

res |> as.data.frame() |>
    dplyr::filter(!is.na(padj)) |>
    dplyr::arrange(padj) |>
    head(5)
```

# Score candidate regulators

Load the demonstration reference database and map the query onto it. As in
the bulk workflow, `prepareQuery()` maps gene identifiers to Entrez and drops
genes with a missing adjusted p-value, which DESeq2 assigns to genes it
filtered out before testing.

```{r prepare-query, message=TRUE}
demo_expression_db <- getReferenceDatabase(demo = TRUE)

query <- prepareQuery(res, reference = demo_expression_db)
head(query)
```

The stimulated CD14+ monocytes carry a continuous fold-change signature, so
we score it with `cosine`.

```{r score}
regulatorResults <- calcEnrichment(
    demo_expression_db,
    expressionMatrix = query,
    method = "cosine"
)

hit_cols <- c("pert_symbol", "pert_type", "entrez_gene_id", "score", "zscore")
topRegulators(regulatorResults, n = 10)[, hit_cols]
```

# Filtering hits that cannot be human regulators

The reference collection is assembled from many public perturbation studies,
and the sample annotation was largely produced by an LLM-assisted pipeline.
That pipeline has erroneously admitted some studies that overexpress viral
proteins in human cells to study host-pathogen interactions, and labelled the
viral protein as the perturbed target. These signatures make biological sense
as top hits for an antiviral interferon response, but the perturbed factor is
not a human gene and so cannot be an upstream regulator in the intended sense.

We do not have to spot these by eye. Because the annotation could not assign a
human Entrez ID to a viral protein, the `entrez_gene_id` column is `NA` for
exactly these cases, so we can filter them out explicitly:

```{r filter-nonhuman}
scored <- regulatorTable(regulatorResults)

nonhuman <- scored[is.na(scored$entrez_gene_id), ]
nonhuman[, c("pert_symbol", "pert_type", "score", "zscore")]
```

Here the two strongest positive hits are the adenoviral protein *E4ORF3* and
the herpesviral protein *ICP0*, both known modulators of innate immune
signalling. Keeping only the rows with a human Entrez ID leaves the top
*human* candidate regulators:

```{r human-only}
human_hits <- scored[!is.na(scored$entrez_gene_id), ]
human_hits <- human_hits[order(-abs(human_hits$zscore)), ]
head(human_hits[, c("pert_symbol", "pert_type", "score", "zscore")])
```

The strongest human hits are the DNA-sensing and interferon machinery:
*STING1* (cGAS-STING pathway), *IFI16*, and *IRF1*. Their reference
signatures are knockouts or knockdowns whose scores are *negative*
(reversing): removing an antiviral regulator produces the opposite of the
interferon response, exactly the relationship expected if that regulator is
required for the response.

We can filter at the PerturbMatch object level also:
```{r filter-perturbMatch}
keep <- !is.na(rowData(regulatorResults)$entrez_gene_id)
regulatorResults <- regulatorResults[keep, ]
```


# Visualise the results

```{r plot-similarity, fig.width=7}
plotSimilarity(regulatorResults, topN = 20)
```

Setting `inferActivity = TRUE` flips the sign of the knockdown/knockout
signatures, so the antiviral regulators move to the mimicking side: the data
are consistent with interferon stimulation *increasing* the activity of
*STING1*, *IFI16*, and *IRF1*.

```{r plot-similarity-activity, fig.width=7}
plotSimilarity(regulatorResults, topN = 20, inferActivity = TRUE)
```

# Cell-level scoring with singscore

perturbMatch ranks regulators from the aggregated pseudobulk signature,
which collapses every cell in a sample to a single value. This
hides any variation between cells.
`r Biocpkg("singscore")` scores each individual cell against a gene
signature, so we can bring a candidate regulator back down to the
single-cell level and look at how its activity is distributed across the
population rather than assuming it is uniform.

The reference contains both a *STING1* overexpression and a *STING1* knockout
signature. They sit on opposite sides of the raw perturbMatch score, the
overexpression *mimicking* the interferon response and the knockout
*reversing* it, but the two are concordant once the direction of each
perturbation is taken into account: both say that interferon stimulation
raises *STING1* activity. We use the knockout here because it is the much
stronger, higher-confidence hit.

Because the signature is a knockout, its direction is inverted relative to
STING1 activity. A knockout lowers its target's pathway, so the genes it
pushes *up* mark *loss* of STING1 signalling and the genes it pushes *down*
mark *active* signalling. To score STING1 activity we therefore reverse the
two sets: the knockout's down-genes become the activity up-set and its
up-genes become the down-set. Because we relabelled nothing, these are gene
symbols, matching the row names of the single-cell object.

```{r sting-signature}
sting_col <- "GSE129436_STING1_KO_1"
sting_lfc <- as.numeric(assay(demo_expression_db, "log2fc")[, sting_col])
symbols <- rowData(demo_expression_db)$Symbol

ord <- order(sting_lfc)
n <- 100
ko_down <- symbols[head(ord, n)]   # most decreased by the STING1 knockout
ko_up   <- symbols[tail(ord, n)]   # most increased by the STING1 knockout

# Reverse the knockout to score STING1 *activity*: genes that fall when
# STING1 is knocked out are the ones that mark active STING1 signalling.
sting_up   <- ko_down
sting_down <- ko_up
```

singscore ranks all genes within each cell, so we work on log-normalised
counts and restrict to reasonably detected genes to keep the matrix compact.
The result is one *STING1* activity score per cell.

```{r singscore}
sce <- logNormCounts(sce)
sce <- sce[rowSums(counts(sce) > 0) >= 10, ]

sting_up   <- intersect(sting_up, rownames(sce))
sting_down <- intersect(sting_down, rownames(sce))

ranked <- rankGenes(as.matrix(logcounts(sce)))
sce$STING1_activity <- simpleScore(
    ranked, upSet = sting_up, downSet = sting_down
)$TotalScore
```

To see how that activity is spread across the population, we embed the cells
with a UMAP. We first reduce the data with PCA and then run UMAP on the
principal components.

```{r umap-compute}
set.seed(1)
sce <- runPCA(sce, ncomponents = 20)
sce <- runUMAP(sce, dimred = "PCA")

umap_df <- data.frame(
    reducedDim(sce, "UMAP"),
    STING1_activity = sce$STING1_activity,
    stim = sce$stim
)
colnames(umap_df)[1:2] <- c("UMAP1", "UMAP2")
```

Colouring one point per cell is unreadable with several thousand overlapping
cells, so we bin the UMAP into hexagons and colour each hexagon by the *mean*
*STING1* activity score of the cells in it. A single-hue scale from near-white
to dark blue, with the limits clipped so a few extreme cells cannot dominate
it, makes the gradient across the population clear.

```{r umap-plot, fig.width=9, fig.height=3.8}
p_activity <- ggplot(umap_df, aes(UMAP1, UMAP2, z = STING1_activity)) +
    stat_summary_hex(bins = 50, fun = mean) +
    scale_fill_gradient(
        low = "#EAF3FB", high = "#08519C",
        limits = c(0, 0.5), oob = scales::squish,
        name = "Mean\nSTING1\nactivity"
    ) +
    labs(title = "STING1 activity") +
    theme_bw()

p_condition <- ggplot(umap_df, aes(UMAP1, UMAP2, colour = stim)) +
    geom_point(size = 0.35, alpha = 0.6) +
    scale_colour_manual(values = c(ctrl = "#4C72B0", stim = "#DD8452")) +
    guides(colour = guide_legend(override.aes = list(size = 2, alpha = 1))) +
    labs(title = "Condition") +
    theme_bw()

cowplot::plot_grid(p_activity, p_condition, nrow = 1)
```

The stimulated and control monocytes separate into two clusters, and the
*STING1* activity is raised across the stimulated cluster, confirming at
single-cell resolution the regulator that perturbMatch flagged from the
pseudobulk signature. The binned view also exposes heterogeneity that the
pseudobulk value hides: the score is graded rather than uniform, so some
regions of the stimulated cluster carry a stronger *STING1* programme than
others.

# Session information

```{r session-info}
sessionInfo()
```
