---
title: >
  Matching a differential expression signature to perturbation references
  using perturbMatch
author:
- name: Jamie Soul
  affiliation:
  - Biochemistry, Cell and Systems Biology, Institute of Systems, Molecular
    & Integrative Biology, Faculty of Health and Life Sciences
  email: jamie.soul@liverpool.ac.uk
date: "`r Sys.Date()`"
package: perturbMatch
output:
  BiocStyle::html_document:
    titlecaps: false
    toc: true
    toc_float: true
    number_sections: true
abstract: >
  perturbMatch matches a transcriptomics signature against a reference
  database of single-gene perturbation signatures, asking what
  transcriptional regulators could be driving the gene expression changes
  observed in an experiment. A query
  signature (a differential expression result, or a simple set of
  up/down-regulated genes) is compared against a large reference collection of
  single-gene perturbation signatures using one of several enrichment methods,
  and candidate regulators are ranked by similarity to the query. This vignette
  walks through preparing input data, scoring it against a small demonstration
  reference database, and visualising the results.
vignette: >
  %\VignetteIndexEntry{perturbMatch}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r style, echo=FALSE, results='hide', message=FALSE}
library(BiocStyle)
library(knitr)
opts_chunk$set(error = FALSE, message = FALSE, warning = FALSE)
opts_chunk$set(fig.asp = 1)
```

# Installation

```{r installation, echo=TRUE, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("perturbMatch")

## Or install the development version from GitHub
# install.packages("devtools")
devtools::install_github("soulj/perturbMatch", build_vignettes = TRUE)
```

Gene-identifier mapping depends on `AnnotationDbi` and `org.Hs.eg.db`, and the
full reference databases are retrieved with `ExperimentHub`; all three are
installed automatically alongside perturbMatch. To reproduce this vignette you
will also need the `airway` example dataset and `DESeq2` for the worked
example, and `GO.db` for the regulator GO enrichment:

```{r vignette-deps, echo=TRUE, eval=FALSE}
BiocManager::install(c("airway", "DESeq2", "GO.db"))
```

# Background

## Overview

perturbMatch matches a transcriptomics signature against a reference database
of single-gene perturbation signatures, asking what transcriptional regulators
could be driving the gene expression changes observed in an experiment.

## Reference signatures

perturbMatch provides a large collection of single-gene perturbation
signatures derived from publicly available RNA-seq datasets. Each signature
captures the transcriptional response to overexpression, knockdown, or
knockout of a single gene.

The signatures are computed from 4,453 public RNA-seq series in NCBI GEO,
covering 3,571 knockdowns, 1,779 knockouts and 1,452 overexpressions over a
common set of 39,376 NCBI genes. Series were selected on the basis of detected
single gene perturbations and availability of pre-processed count data.
The GEO series behind each signature is recorded in its `sig_id` 
and `accession`, so any hit can be traced back to its source experiment.

Assigning a perturbation type, a target gene and case/control sample groups to
each series was automated with an LLM-assisted curation pipeline. The model
used is recorded per signature in `colData(reference)$model`. Automated
curation is not error-free, so [Verifying hits before interpretation] shows how
to check a result before acting on it.

The reference data are distributed as the `perturbMatchData` ExperimentHub
package and archived at Zenodo under
[doi:10.5281/zenodo.21792179](https://doi.org/10.5281/zenodo.21792179),
released under a CC-BY-4.0 licence.

## How it works

perturbMatch compares your experimental gene expression signature against
each reference signature using one of several implemented enrichment methods:
`SignedJaccard`, `GSEA`, `cosine` or `cosineExtreme`. These are described
in [Scoring methods available]. Regulators whose perturbation signatures
closely match your query are ranked as likely upstream drivers.

# Preparing input data

We start from example RNA-seq data, which is analysed with DESeq2 to
identify differentially expressed genes. The differential expression
results can be passed directly to perturbMatch to be screened against a
large collection of potential regulators.

perturbMatch accepts input as a \pkg{DESeq2} `DESeqResults` object, a
\pkg{limma} `topTable()` data frame, or any data frame with gene identifiers
and a log2 fold-change column. `prepareQuery()` (below) converts any of
these into the two-column form `calcEnrichment()` expects, and also
maps the gene identifiers to Entrez IDs used by the reference
databases.

Correlation-based methods (see [Scoring methods available])
require the full set of results from a differential expression analysis,
i.e. all genes measured in the experiment, not just the statistically
significant subset.

The example below uses the `airway` dataset, which measures the gene
expression response to dexamethasone treatment in airway smooth muscle
cells. The `airway` dataset is loaded into the session with `data(airway)`
after `library(airway)`. We perform differential expression analysis on this
data with DESeq2 to create an example gene expression response for querying.

```{r input-data}
library(perturbMatch)
library(SummarizedExperiment)
library(BiocParallel)
library(airway)
library(DESeq2)

# Load example data
data(airway)

# Calculate differential expression using DESeq2
dds_airway <- DESeqDataSet(airway, design = ~ cell + dex) |>
    DESeq() |>
    results(contrast = c("dex", "trt", "untrt"))
```

We now have a `DESeqResults` object, `dds_airway`, containing the log2 fold
changes and adjusted p-values for every gene measured in the experiment. This
is the input we will pass to perturbMatch.

```{r explore-dds}
head(dds_airway)
```


## Load reference expression data

Select and load the reference gene expression data to compare against your
experiment with `getReferenceDatabase()`. The reference database is
returned as a `SummarizedExperiment` object.

For the purposes of this vignette we use a small bundled demonstration
database of 21 signatures (`demo = TRUE`). The demonstration database
should not be used for any real analysis: it is a small subset of the
full reference database, chosen to illustrate both this bulk example and the
single-cell workflow, and provided for illustration only.

```{r load-data, message=TRUE}
demo_expression_db <- getReferenceDatabase(demo = TRUE)
```

### Choosing a full reference database

The full reference databases are retrieved with `getReferenceDatabase()`
(`demo = FALSE` is the default) from ExperimentHub, where they are registered
by the `perturbMatchData` package. They are between 0.7-3.1 GB, but only need
to be downloaded once: ExperimentHub caches them, and `metadata()` on the
returned object records which source they came from. Several builds of the
reference are available through the `type` and `batchCorrection` arguments:

* **`type = "chrdir"`** gives signatures summarised with the
    characteristic-direction method. This is the default.
* **`type = "limma"`** gives signatures derived from a `limma` differential
    expression analysis useful for producing volcano plots of the signatures.

Each type comes in two variants, selected with `batchCorrection`: the
uncorrected signatures (`batchCorrection = FALSE`, the default) and signatures
computed with study or batch effects removed (`batchCorrection = TRUE`).

```{r reference-options, eval=FALSE}
# Characteristic-direction reference, without batch-effect correction (default)
ref_chrdir <- getReferenceDatabase(type = "chrdir")

# Characteristic-direction reference, with batch-effect correction
ref_chrdir_bc <- getReferenceDatabase(type = "chrdir", batchCorrection = TRUE)

# limma reference, without batch-effect correction (default)
ref_limma <- getReferenceDatabase(type = "limma")

# limma reference, with batch-effect correction
ref_limma_sva <- getReferenceDatabase(type = "limma", batchCorrection = TRUE)
```

`chrdir` references store a single characteristic-direction score, while `limma`
references store both the log2 fold change (`log2fc`) and a combined *pi* score
(`log2FC * -log10(padj)`, the default), which weights each gene's fold
change by its statistical significance. `calcEnrichment()`'s `refMetric`
argument selects which of a reference database's stored metrics to score
against (for example `refMetric = "log2fc"` on a `limma` reference). The
bundled demonstration database used throughout this vignette is a small
`limma` subset, so its default metric is *pi*.

Data are not loaded into R until required, keeping RAM usage low. The first
few rows and columns of the assay can be inspected:

```{r explore-db}
assay(demo_expression_db)[1:3, 1:3]
```

Metadata for each perturbation experiment are stored in the `colData` slot:

```{r explore-coldata}
coldata_cols <- c(
    "sig_id", "pert_symbol", "pert_type", "cell_line", "cell_line_efo_label"
)
head(colData(demo_expression_db)[, coldata_cols], 3)
```

Any `SummarizedExperiment` object containing a measure of differential
expression can be used as a reference database for perturbMatch, provided
`colData` includes at least the `pert_symbol`, `sig_id`, and `pert_type`
columns identifying each experiment. The remaining columns in Table 1 are
annotation that the packaged references also provide: provenance of the
source experiment (`accession`, `control_samples`, `case_samples`,
`technology`), how the signature was curated (`model`), the biological
context it was generated in (`cell_line`, `cell_line_efo_id`,
`cell_line_efo_label`), and per-signature quality information
(`entrez_gene_id`, `target_gene_percentile`).

```{r table-coldata, echo=FALSE}
descriptions <- c(
    sig_id = paste(
        "Unique identifier for the perturbation signature.",
        "Matches the signature matrix column names"
    ),
    pert_symbol = "Official gene symbol of the perturbed gene target",
    pert_id = "Internal ID combining the symbol and perturbation type",
    pert_type = paste(
        "Type of perturbation: `KD` (knockdown), `KO` (knockout)",
        "or `OE` (overexpression)"
    ),
    accession = "NCBI GEO series accession for the source study",
    control_samples = paste(
        "Comma-separated list of raw sample IDs used as the control"
    ),
    case_samples = "Comma-separated list of raw sample IDs used as the case",
    technology = paste(
        "Profiling platform used to generate the source data,",
        "e.g. `RNA-seq`"
    ),
    cell_line = paste(
        "Cell line or primary cell type the perturbation was performed in,",
        "as free text reported by the source study, e.g. `T47D`,",
        "`iPSC-derived Macrophages`"
    ),
    model = paste(
        "Curation approach used to annotate the signature: `curated` by",
        "hand, or the LLM that extracted the annotation"
    ),
    entrez_gene_id = paste(
        "NCBI Entrez ID of the perturbed gene. `NA` when the target has no",
        "human Entrez entry, e.g. a viral gene. Use it to filter out hits",
        "that cannot be human upstream regulators"
    ),
    cell_line_efo_id = paste(
        "Experimental Factor Ontology (EFO) identifier matched to",
        "`cell_line`, e.g. `EFO:0003043`. `NA` where no confident match",
        "was found, so not every signature is annotated"
    ),
    cell_line_efo_label = paste(
        "EFO label for `cell_line_efo_id`, giving a standardised name for",
        "the free-text `cell_line` value (`NCI-H1299` for `H1299`).",
        "Use it to group or select signatures by cell context"
    ),
    target_gene_percentile = paste(
        "Percentile position of the perturbed gene within its own",
        "signature, from most down-regulated (0) to most up-regulated",
        "(100). It is a quality check on whether the perturbation worked"
    )
)

# Ordered by the colData columns themselves, so the table cannot drift from
# the object it documents.
schema_cols <- colnames(colData(demo_expression_db))
stopifnot(setequal(schema_cols, names(descriptions)))

schema_df <- data.frame(
    `Column Name` = schema_cols,
    `Description` = unname(descriptions[schema_cols]),
    check.names = FALSE
)

knitr::kable(
    schema_df,
    row.names = FALSE,
    caption = "Table 1: Reference database colData columns"
)
```

## Map the query to the reference

The `airway` DESeq2 result uses Ensembl gene IDs, but the
reference database uses Entrez IDs. `prepareQuery()` maps between the two,
reporting how much of the query mapped. Passing `reference` additionally
reports how much of the mapped query is actually present in the reference,
a check that the gene IDs line up before scoring.

By default `prepareQuery()` also drops genes with a missing adjusted
p-value. `DESeq2` leaves `padj` as `NA` for genes it filters out before
testing (typically because they are too lowly expressed to be informative),
and these carry no reliable signal for signature matching. Set
`dropLowExpressed = FALSE` to keep them.

```{r prepare-query, message=TRUE}
final_res <- prepareQuery(dds_airway, reference = demo_expression_db)
head(final_res)
```

## Starting from a table or CSV file

`prepareQuery()` also accepts any `data.frame`, so you can read a file
with `read.csv()` (or `read.delim()` for tab-separated files) to be used
as input.

The table needs:

* a gene identifier for each row, either in the row names or in a
    column. Ensembl, Entrez or gene symbols are all fine.
    `prepareQuery()` infers the type and maps to Entrez.
* a log2 fold-change column named `log2FoldChange`, `logFC`, or
    `log2FC`.
* *(optional)* an adjusted p-value column named `padj` or `adj.P.Val`,
    which enables the low-expression filtering described above.

Columns are matched by name, so the fold-change and p-value columns can sit
anywhere in the table.

Anything `prepareQuery()` does not recognise such as raw p-values, base
means, average expression is ignored, and the returned query always has the
same `gene`/`log2FC`/`padj` columns.

To keep this example self-contained we first write the airway result out to
a temporary CSV file, deliberately putting the columns in an awkward order:

```{r csv-write}
de_table <- as.data.frame(dds_airway)
de_table <- data.frame(
    baseMean = de_table$baseMean,       # extra column, ignored
    padj     = de_table$padj,           # optional adjusted p-value
    gene_id  = rownames(de_table),      # gene ID, not in the first column
    pvalue   = de_table$pvalue,         # extra column, ignored
    log2FC   = de_table$log2FoldChange  # fold-change column
)

csv_path <- tempfile(fileext = ".csv")
write.csv(de_table, csv_path, row.names = FALSE)
```

Reading it back in and preparing it then gives the same query as before:

```{r csv-read, message=TRUE}
my_de <- read.csv(csv_path)
head(my_de)

query_from_csv <- prepareQuery(my_de, reference = demo_expression_db)
head(query_from_csv)
```

# Calculating regulator scores

With the query mapped to the reference, we can score every reference
signature against it. Here we score the full ranked expression matrix using
the `cosine` method, which is appropriate for continuous fold-change data
(see [Scoring methods available] for alternatives suited to gene-set input).

```{r scoring}
regulatorResults <- calcEnrichment(
    demo_expression_db,
    expressionMatrix = final_res,
    method = "cosine"
)
```

Results are stored in a `SummarizedExperiment`-like `PerturbMatch` object,
with enrichment scores in the `enrichment` assay and perturbation experiment
details in `rowData`. Printing it gives a quick summary of both:

```{r explore-perturbMatch-show}
regulatorResults
```

Perturbation details for each reference signature carry through into
`rowData` (showing a few key columns here):

```{r explore-perturbMatch-rowdata}
head(rowData(regulatorResults)[, c("pert_symbol", "pert_type", "cell_line")])
```

The up-/down-regulated and background gene sets actually used for scoring
are recorded on the result and can be retrieved with the corresponding
accessors.

```{r explore-perturbMatch-genesets}
upRegulated(regulatorResults) |> head()
downRegulated(regulatorResults) |> head()
background(regulatorResults) |> head()
```

View the top hits ranked by absolute score with `topRegulators()`:

```{r results}
cosine_cols <- c("pert_symbol", "pert_type", "score", "zscore")
topRegulators(regulatorResults, n = 5)[, cosine_cols]
```

## Checking a known target

The airway data measure the response to dexamethasone, a glucocorticoid
that acts directly through the glucocorticoid receptor, *NR3C1*. *NR3C1* is
therefore the expected upstream regulator, and the small demo database was
selected to include a *NR3C1* knockdown signature. We rank the reference
signatures by absolute score and look up where *NR3C1* is ranked:

```{r known-target}
ranked <- topRegulators(regulatorResults, n = nrow(regulatorResults))
nr3c1_rank <- match("NR3C1", ranked$pert_symbol)
nr3c1_rank

ranked[nr3c1_rank, cosine_cols]
```

The *NR3C1* knockdown ranks near the top (rank `r nr3c1_rank` of
`r nrow(regulatorResults)`). Its raw score is *negative*: dexamethasone
*activates* NR3C1 while the reference signature *knocks it down*, so the two
run in opposite directions. The [Visualise the results] section below turns
this into an activity-level statement with `inferActivity = TRUE`.

> **A note on `zscore` and reference set size.** `zscore` standardises each
> method's raw score across the reference signatures actually scored in
> that call, so its scale depends on the size and diversity of that
> background set. This demo database has only `r nrow(regulatorResults)`
> signatures, nowhere near enough for these particular z-scores to be
> reliable. With the full reference database (thousands of signatures) the
> same scores become much more informative.

## Using an explicit gene set

Instead of a ranked matrix, `calcEnrichment()` also accepts an explicit
query as `upRegulated`/`downRegulated`/`background` character vectors of
gene IDs. Use this format when you only have a fixed set of significant
genes rather than a full ranked differential expression result. If you
do have an `expressionMatrix`, `SignedJaccard` and `GSEA` will derive an
up/down gene set from it for you (as `cosine`/`cosineExtreme` always
require), so there is no need to manually build the sets in this case. Here
we derive a significant gene set from the same DESeq2 result used above
(adjusted p-value < 0.05), and score it with `SignedJaccard`:

```{r geneset-input}
sig_genes <- final_res[!is.na(final_res$padj) & final_res$padj < 0.05, ]

up_genes <- sig_genes$gene[sig_genes$log2FC > 0]
down_genes <- sig_genes$gene[sig_genes$log2FC < 0]
background_genes <- final_res$gene

results_signedjaccard <- calcEnrichment(
    demo_expression_db,
    upRegulated = up_genes,
    downRegulated = down_genes,
    background = background_genes,
    method = "SignedJaccard"
)

result_cols <- c("pert_symbol", "pert_type", "score", "overlap", "zscore")
topRegulators(results_signedjaccard, n = 5)[, result_cols]
```

# Visualise the results

`plotSimilarity()` displays the top positive (mimicking) and negative
(reversing) signatures as a lollipop plot:

```{r plot-similarity, fig.width=7}
plotSimilarity(regulatorResults, topN = 20)
```

The fill colour shows the sign of the similarity score, orange for
signatures that *mimic* the query (positive score) and blue for those that
*reverse* it (negative score). The point shape shows the direction of
the underlying perturbation: up-pointing triangles for gain-of-function
overexpression (`OE`) and down-pointing triangles for loss-of-function
knockdown/knockout (`KD`/`KO`). Keeping these separate lets you read the
perturbation type together with the sign of the score. A *reversing* (blue)
*knockdown* (down-triangle) signature, for instance, implies the query
behaves as though that gene's activity were *increased*.

Set `inferActivity = TRUE` to flip the sign of scores for knockdown/
knockout perturbations, so the plot reflects inferred gene *activity*
rather than raw transcriptional similarity. A knockdown/knockout signature
that is *anti*-correlated with the query (a negative raw score) implies
the query behaves as though that gene's activity had *increased*, the
opposite of what a negative score would mean for an overexpression
signature.

For example, `AFF4_KD` has a raw cosine z-score of about
`r round(regulatorTable(regulatorResults)["GSE197354_AFF4_KD_1", "zscore"], 2)`,
placing it on the "Reverse (Negative)" side of the plot above: the knockdown
signature runs opposite to the dexamethasone response. Because it is a
knockdown, `inferActivity = TRUE` flips its sign to about
`r round(-regulatorTable(regulatorResults)["GSE197354_AFF4_KD_1", "zscore"], 2)`
below, moving it to the "Mimic (Positive)" side: the data are consistent
with dexamethasone *increasing* AFF4 activity, since suppressing AFF4
produces roughly the opposite expression pattern to the treatment.

```{r plot-similarity-activity, fig.width=7}
plotSimilarity(regulatorResults, topN = 20, inferActivity = TRUE)
```

## Inspecting a single reference signature

Before trusting a top hit it is worth looking at the reference signature
itself: did the perturbed gene actually move in the source data, and does
the overall signature look like a real, well-behaved differential
expression result? `plotVolcano()` draws a volcano plot (log2 fold change
against `-log10` FDR) for any single column of a reference database that
stores per-gene fold changes and p-values (i.e. the `limma` references and
the demo database used here).

Continuing the dexamethasone example, we plot the *NR3C1* knockdown
signature and highlight the perturbed gene itself with `selectedPoints`.
The reference is indexed by Entrez ID, so we relabel the rows with gene
symbols first to make the plot readable.

```{r plot-volcano, fig.width=7}
nr3c1_signature <- demo_expression_db
rownames(nr3c1_signature) <- rowData(nr3c1_signature)$Symbol

plotVolcano(
    nr3c1_signature,
    signature         = "GSE200312_NR3C1_KD_1",
    effectAssay       = "log2fc",
    pvalueAssay       = "padj",
    selectedPoints    = "NR3C1",
    truncateInfPvalue = FALSE
)
```

*NR3C1* itself sits well down the left-hand (down-regulated) side of its own
knockdown signature, confirming the intended perturbation took effect.

## Where the perturbed gene ranks: `target_gene_percentile`

The volcano plot gives the differential expression of the perturbed gene one
signature at a time. `colData()` records the same information for every
reference signature in a single column, `target_gene_percentile`: the
percentile position of the perturbed gene within its own signature, ranked
from most down-regulated (0) to most up-regulated (100).

The direction of the perturbation sets the expectation. A knockout or
knockdown (`KO`/`KD`) suppresses its target, so the perturbed gene should
sit near the bottom of its signature and score a *low* percentile. An
overexpression (`OE`) drives its target up, so it should sit near the top
and score a *high* percentile. A value that contradicts the perturbation
type is a potential warning sign that the signature is mislabelled or that the
perturbation did not take effect.

```{r target-percentile}
pct_cols <- c("pert_symbol", "pert_type", "target_gene_percentile")
pct <- as.data.frame(colData(demo_expression_db)[, pct_cols])
pct <- pct[order(pct$target_gene_percentile), ]
pct
```

The *NR3C1* knockdown from the volcano plot above illustrates the low end:

```{r target-percentile-worked}
nr3c1_pct <- pct[pct$pert_symbol == "NR3C1", ]
nr3c1_pct
```

*NR3C1* sits in the bottom 1% of its own knockdown signature
(`target_gene_percentile` of about
`r round(nr3c1_pct$target_gene_percentile, 1)`), near the bottom as a `KD`
should be, and matching where it appeared on the left of the volcano plot.
At the other extreme, the overexpression signatures (`OE`) place their
targets close to the 100th percentile.

Note the *CCDC88A* is annotated as a knockout, so its target should sit near
the bottom, yet it lands around the
`r round(pct$target_gene_percentile[pct$pert_symbol == "CCDC88A"])`th
percentile, the middle of its own signature:

```{r target-percentile-suspect}
pct[pct$pert_symbol == "CCDC88A", ]
```

Signatures such as this should be treated with caution as the perturbation may
have been weak or incomplete, or the signature may be mislabelled. This is
why [Verifying hits before interpretation] is important before any candidate
is followed up.

### Filtering on the target percentile

`target_gene_percentile` is an ordinary `colData` column, so signatures can
be dropped with standard `SummarizedExperiment` subsetting before scoring.
Here we require knockdowns to place their target in the bottom quarter of
their own signature, leaving the other perturbation types untouched:

```{r target-percentile-filter}
target_pct <- colData(demo_expression_db)$target_gene_percentile
pert_types <- colData(demo_expression_db)$pert_type

drop <- pert_types == "KD" & !is.na(target_pct) & target_pct > 25
filtered_db <- demo_expression_db[, !drop]

c(all = ncol(demo_expression_db), filtered = ncol(filtered_db))
```

`filtered_db` is a reference database and can be passed
straight to `calcEnrichment()`.

Note that a CRISPR knockout
(`KO`) won't necessarily reduce the target gene expression 
(only the target protein), so non-low percentile is not by
itself evidence that the perturbation failed.

# Functional enrichment of the regulator hits

Individual top hits are easier to interpret as a group when you ask what
biological processes the mimicking and reversing regulators have in common.
`regulatorGSEA()` does this by ranking the reference regulators by their
similarity score and performing gene set enrichment (via `fgsea`) over Gene
Ontology terms.

A regulator gene usually appears as several signatures (different cell
lines, knockdown versus overexpression). `regulatorGSEA()` collapses these
to one value per gene by flipping the sign of knockdown/knockout scores
so everything is on a common inferred-activity scale (as in
`plotSimilarity(inferActivity = TRUE)`), then aggregating each gene's
signatures (`aggregate = "extreme"` by default, keeping the single strongest
signature, or `"mean"` to average them).

Gene sets are built from `org.Hs.eg.db` and readable GO term names are added
from `GO.db` when it is installed.

```{r regulator-gsea}
go <- regulatorGSEA(results_signedjaccard, ontology = "BP", minSize = 1)
head(go[, c("term", "NES", "padj", "size", "leadingEdge")])
```

`plotRegulatorGSEA()` draws a dot plot from that table, placing each term by
its `NES`:

```{r regulator-gsea-plot, fig.width=7, fig.height=5}
plotRegulatorGSEA(go, n = 10, pvalueCutoff = 1)
```

On this 21-signature demo database the enrichment is underpowered and none of
the terms are individually significant, so `pvalueCutoff = 1` is used purely
to render an illustrative plot and `minSize` is dropped to 1. On a full
reference (thousands of regulators) keep the defaults (`minSize = 15`,
`pvalueCutoff = 0.1`), where the enrichment becomes meaningful and the terms
summarise the biology shared by the top regulators.

# Scoring methods available

```{r table-methods, echo=FALSE}
methods_data <- data.frame(
    Method = c("cosine", "cosineExtreme", "GSEA", "SignedJaccard"),
    `Input Type` = c(
        "Continuous", "Rank/set-based", "Continuous/ranked", "Set-based"
    ),
    Description = c(
        paste(
            "Classic cosine similarity between the full continuous log2FC",
            "vectors of the query and reference."
        ),
        paste(
            "Cosine similarity restricted to the top N most extreme genes",
            "of the reference and matching query features."
        ),
        paste(
            "Gene Set Enrichment Analysis. It evaluates whether the query",
            "gene set or ranked list is enriched at the top or bottom of the",
            "reference signature distribution."
        ),
        paste(
            "Intersection-over-union (Jaccard index) for gene sets, taking",
            "into account the direction (sign) of the log2FC regulation."
        )
    ),
    check.names = FALSE
)

knitr::kable(
    methods_data,
    row.names = FALSE,
    caption = "Table 2: Overview of reference expression scoring methods"
)
```

# Comparing scoring methods

The methods above rank the same reference signatures on different notions
of similarity, so they broadly agree but need not identify identical top
hits. Having already scored the airway/dexamethasone query with `cosine`
and `SignedJaccard` above, we complete the comparison by also running
`GSEA` (which, like `SignedJaccard`, takes an explicit gene set) and
`cosineExtreme` (which, like `cosine`, takes the ranked matrix):

```{r compare-methods-run}
results_gsea <- calcEnrichment(
    demo_expression_db,
    upRegulated = up_genes,
    downRegulated = down_genes,
    background = background_genes,
    method = "GSEA"
)

results_cosineextreme <- calcEnrichment(
    demo_expression_db,
    expressionMatrix = final_res,
    method = "cosineExtreme"
)
```

Aligning the four result sets by signature and correlating their z-scores
shows how much they overlap. Methods within the same family, the continuous
`cosine`/`cosineExtreme` pair and the set-based `SignedJaccard`/`GSEA` pair,
tend to track each other closely, while agreement *across* the two families
is typically weaker, because they measure genuinely different things:

```{r compare-methods-cor}
sig_ids <- rownames(regulatorResults)

score_compare <- data.frame(
    cosine        = regulatorTable(regulatorResults)$zscore,
    cosineExtreme = regulatorTable(results_cosineextreme)[sig_ids, "zscore"],
    SignedJaccard = regulatorTable(results_signedjaccard)[sig_ids, "zscore"],
    GSEA          = regulatorTable(results_gsea)[sig_ids, "zscore"]
)

round(cor(score_compare), 2)
```

On this small demo database the correlations are good, but on a full-size
reference (thousands of signatures) `cosine` and `SignedJaccard` can produce
disconcordant rankings. `cosine` asks whether the whole fold-change profile
of the query is aligned with the reference, so every gene contributes in
proportion to its magnitude. `SignedJaccard`/`GSEA` first reduce each
signature to a discrete up/down gene set and then ask how much those sets
overlap in the same direction, so they are driven by which genes cross the
threshold rather than by how large their changes are. A regulator can
therefore rank highly under one family and modestly under the other
depending on whether its signal lives in a broad continuous shift or a
concentrated set of strongly moved genes.

In practice, `cosine`/`cosineExtreme` are a reasonable default when a full
ranked signature is available, while `SignedJaccard`/`GSEA` are useful when
only a discrete up/down gene set can be defined, or when a set-based
interpretation (shared genes, rather than correlated magnitudes) is
preferred.

## Interpreting the score, zscore, and overlap columns

Every `perturbMatch` result carries at least two columns beyond the method's
own score: `zscore` , the number of standard deviations a signature's score
sits from the mean across the reference queried (used for ranking with
`topRegulators()` and for `plotSimilarity()`) and, for `SignedJaccard`,
`overlap`, the number of query genes that overlapped the reference gene set
at all (in either direction). A large `|zscore|` backed by a small `overlap`
is weaker evidence than the same `|zscore|` backed by a large one, since it
is based on fewer shared genes:

```{r overlap-filter}
reliable_hits <- regulatorTable(results_signedjaccard)
reliable_hits <- reliable_hits[reliable_hits$overlap >= 50, ]
reliable_hits <- reliable_hits[order(-abs(reliable_hits$zscore)), ]

head(reliable_hits[, result_cols])
```

# Tuning topN

`SignedJaccard` and `GSEA` need reference gene sets, not just ranked
values, so each reference signature's continuous scores are first
discretised into up-/down-regulated sets. By default this is done with
`topN`: the `topN` highest- and lowest-ranked reference genes become that
signature's up-/down-regulated sets.


```{r tune-topn}
sj_small <- calcEnrichment(
    demo_expression_db, upRegulated = up_genes, downRegulated = down_genes,
    background = background_genes, method = "SignedJaccard", topN = 100
)
sj_large <- calcEnrichment(
    demo_expression_db, upRegulated = up_genes, downRegulated = down_genes,
    background = background_genes, method = "SignedJaccard", topN = 1000
)
```

*NR3C1* is the expected regulator of the dexamethasone target, so we find
its rank under each setting to see the effect these settings have:

```{r tune-topn-rank}
nr3c1_rank_at <- function(res) {
    ranked <- topRegulators(res, n = nrow(res))
    match("NR3C1", ranked$pert_symbol)
}

c(
    topN_100  = nr3c1_rank_at(sj_small),
    topN_1000 = nr3c1_rank_at(sj_large)
)
```

The *NR3C1* knockdown ranks `r nr3c1_rank_at(sj_small)` out of
`r nrow(sj_small)` with `topN = 100` but slips to
`r nr3c1_rank_at(sj_large)` with `topN = 1000`: widening the reference gene
sets dilutes the overlap that placed it near the top. Comparing a few `topN`
settings this way is a reasonable check that a finding is not strongly
dependent on the cutoff chosen.


# Parallelisation

Scoring the full reference database (thousands of signatures) is dominated
by reading and decompressing parts of the reference database, and this work
parallelises well. `calcEnrichment()` accepts a `BiocParallel` back-end via its
`BPPARAM` argument. Supplying a multi-core parameter shares the reference
signatures across workers. On a typical multi-core machine this runs roughly
3 to 4 times faster than the default.

```{r parallel, eval=FALSE}
library(BiocParallel)

regulatorResults <- calcEnrichment(
    ref_limma,
    expressionMatrix = final_res,
    method  = "cosine"
)
```

Work is split into column wise chunks whose size is controlled by `chunkSize`.
By default (`chunkSize = NULL`) `calcEnrichment()` picks a chunk size
automatically from the number of reference signatures and the number of
workers in `BPPARAM`. Because each worker holds its own block of the
reference in memory, more workers means higher peak RAM usage.

# Verifying hits before interpretation

The reference database is assembled from public datasets whose metadata
(which samples are cases versus controls, and which gene was perturbed) is
**majority LLM-annotated**. This automated curation makes the collection
broad and has been benchmarked against hand curated data, but it is not
error-free: mislabelled case/control groups or incorrect perturbation targets
are likely to occur.

**Before interpreting or following up a top hit, manually verify it.** For
each candidate regulator, check the original data source (the `accession`,
`control_samples`, and `case_samples` columns in `rowData()`) to confirm
that the case/control assignment is correct and that the annotated perturbed
gene is the one actually manipulated.

**Cell context matters.** Each reference signature reflects the perturbation
of a gene in a particular cell line or cell type (the `cell_line` column of
`rowData()`), and a regulator's downstream program can differ substantially
between contexts. A genuine driver may therefore score only modestly if its
available reference signatures come from cell types unrelated to your
experiment, while signatures generated in a matching context tend to rank
more strongly. Consider how closely its `cell_line` resembles your own
system, and potentially give more weight to regulators supported by several
signatures across relevant contexts than to those resting on a single
signature from an unrelated cell type. Because `cell_line` is free text
taken from the source study, the same cell line can appear under several
spellings. Where an ontology match was available, the `cell_line_efo_id` and
`cell_line_efo_label` columns give a standardised Experimental Factor
Ontology name for it, which is the more reliable field to group or filter
signatures on.

# Saving results to a file

Any of the tables above is an ordinary `data.frame`, so it can be written
out with `write.table()` for use in Excel or a downstream script. To save the
ranked regulators as a tab-separated (TSV) file:

```{r save-tsv, eval=FALSE}
write.table(
    topRegulators(regulatorResults, n = nrow(regulatorResults)),
    file = "perturbmatch_regulators.tsv",
    sep = "\t", quote = FALSE, row.names = FALSE
)
```

# Related packages

perturbMatch screens a query against a curated collection of single-gene
genetic perturbation RNA-seq signatures to nominate upstream regulators.
Several other Bioconductor packages address related questions (Table 3). In
particular, connectivity against drug/compound signatures (from the L1000
platform) is covered by `signatureSearch` rather than perturbMatch.

```{r table-related, echo=FALSE}
related <- data.frame(
    Package = c(
        "signatureSearch", "decoupleR / dorothea", "viper / RTN", "gCMAP"
    ),
    Focus = c(
        "Drug/compound connectivity (LINCS L1000, CMap) for drug repurposing",
        "Transcription-factor activity from curated regulatory networks",
        "Master-regulator analysis over inferred regulatory networks",
        "General framework for building and querying connectivity maps"
    ),
    `Contrast with perturbMatch` = c(
        "perturbMatch queries genetic perturbations, not chemical ones",
        "Knowledge/regulon-based, whereas perturbMatch is data-driven",
        "Network-inferred regulators, not measured perturbation signatures",
        paste(
            "perturbMatch includes a curated perturbation reference",
            "and scoring methods"
        )
    ),
    check.names = FALSE
)

knitr::kable(
    related,
    row.names = FALSE,
    caption = "Table 3: Related Bioconductor packages"
)
```

# Session information

This vignette was compiled using:

```{r session-info, echo=TRUE, message=FALSE}
sessionInfo()
```

# Acknowledgements

This package was developed at the University of Liverpool.
